Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-16 talk/iwtc11/paper.tex
fixes

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

The operations in the trace are indented
corresponding to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned a value exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \lstinline{y} and \lstinline{res} in the while-loop of
the original function.

@@

Loop peeling is achieved by appending a copy of the traced iteration at
the end of the loop. See Figure~\ref{fig:overview}
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore the net effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to

@@

The guard on $p_5$ on line 17 of Figure~\ref{fig:unopt-trace} can be
removed since $p_5$ is allocated on line 10 with a known class. The
guard on $p_0$ on line 20 can be removed since it is identical to the
guard on line 6.

@@

The issue at hand is to keep the peeled loop a proper
trace. Consider the \lstinline{get} operation on line 19 of
Figure~\ref{fig:unopt-trace}. The result of this operation can be
deduced to be $i_4$ from the \lstinline{set} operation on line
12. Also, the result of the \lstinline{get} operation on line 22 can
be deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 19 and 22 from the trace and
replace $i_6$ with $i_4$ and $i_7$ with $i_3$.

@@

In general what is needed is for the heap optimizer is to keep track of
which variables from the preamble it reuses in the peeled loop.
It has to construct a vector of such variables $H$ which
can be used to update the input and jump arguments using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to:

@@

@@

Consider again the original unoptimized trace of
Figure~\ref{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as virtual. This means
that it refers to an virtual object that was not yet been
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the virtual
reference $p_5$. This can be achieved by exploding $p_5$ into it's
attributes. In this case there is only one attribute and it's value is
$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all registered attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non-virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non-
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
and the arguments of the \lstinline{jump} operation of the peeled loop,
$K$, constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to mach the first. This means
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are virtual 
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the first iteration being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?

@@

\subsection{Conclusions}
In this paper we have studied loop invariant code motion during trace
compilation. We claim that loop peeling is a very convenient solution
here since it fits well with other trace optimizations. By peeling of
the first iteration and optimizing the resulting two iteration trace
as a single trace, several standard optimizations can be 
used unchanged. The only interaction needed between the loop peeling
and the other 
optimizations is during the constructing of the jump arguments
connecting the peeled of iteration (the preamble) with the loop body. This
approach 
turns standard optimizations such as redundant guard removal, heap
caching, pure operation reuse and allocation removals into loop
invariant code motion optimizations.

@@

By using several benchmarks we show that the proposed algorithm can
significantly
improve the run time of small loops containing numerical
calculations. 
At least in cases where there are not too many guard
failures. The standard way of handling guards that fail often is to 
trace a bridge from it back to the start of some previously compiled
loop. This is applicable here too. However the bridge will have to end
with a jump to the preamble, which lessens the impact of the 
proposed approach.
In many situations it is however possible to make the bridge
jump to the peeled loop instead. When and how this is possible will be
focus of future work.

>>>>

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

The operations in the trace are indented
corresponding to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned a value exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \lstinline{y} and \lstinline{step} in the while-loop of
the original function.

@@

Loop peeling is achieved by appending an inlined copy of the traced iteration at
the end of itselfe. See Figure~\ref{fig:overview}.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. Looking  at the peeled version of our example in Figure~\ref{fig:peeled-trace} we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore one effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to

@@

The guard on $p_5$ on line 17 of Figure~\ref{fig:peeled-trace} can be
removed since $p_5$ is allocated on line 10 with a known class. The
guard on $p_0$ on line 20 can be removed since it is identical to the
guard on line 6.

Note that the guard on $p_5$ is removed even though $p_5$ is not loop
invariant, which shows that loop invariant code motion is not the only
effect of loop peeling.

@@

The issue at hand is to keep the peeled loop a proper
trace. Consider the \lstinline{get} operation on line 19 of
Figure~\ref{fig:peeled-trace}. The result of this operation can be
deduced to be $i_4$ from the \lstinline{set} operation on line
12. Also, the result of the \lstinline{get} operation on line 22 can
be deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 19 and 22 from the trace and
replace $i_6$ with $i_4$ and $i_7$ with $i_3$.

@@

In general what is needed is for the heap optimizer is to keep track of
which variables from the preamble it reuses in the peeled loop.
It has to construct a vector, $H$,  of such variables which
can be used to update the input and jump arguments using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:peeled-trace} will be optimized to:

@@

Note how the loop invaraint \lstinline{get} on $p_0$ was moved out of
the loop, and how the non loop invariant \lstinline{get} on $p_5$ was
removed entierly.

@@

Consider again the original unoptimized trace of
Figure~\ref{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as virtual. This means
that it refers to an virtual object that has not yet been
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the virtual
reference $p_5$. This can be achieved by exploding $p_5$ into it's
attributes. In this case there is only one attribute and it's value is
$i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all registered attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non-virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non-
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to mach the first jump. This means that
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are virtual 
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the preamble being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?

@@

\subsection{Conclusions}
In this paper we have studied loop invariant code motion during trace
compilation. We claim that loop peeling is a very convenient solution
here since it fits well with other trace optimizations. By peeling of
the first iteration and optimizing the resulting two iteration trace
as a single trace, several standard optimizations can be 
used unchanged. The only interaction needed between the loop peeling
and the other 
optimizations is during the constructing of the jump arguments
connecting the peeled of iteration (the preamble) with the loop body. This
approach 
improves the effect of standard optimizations such as redundant guard removal, heap
caching, common subexpression elimination and allocation removals. The
most prominent effect is that they all become loop
invariant code motion optimizations.

@@

By using several benchmarks we show that the proposed algorithm can
significantly
improve the run time of small loops containing numerical
calculations. 
At least in cases where there are not too many guard
failures. A common way of handling a guard that fails often is to 
trace a bridge from it back to the start of some previously compiled
loop. This is applicable here too. However the bridge will have to end
with a jump to the preamble, which lessens the impact of the 
proposed approach.
In many situations it is however possible to make the bridge
jump to the peeled loop instead. When and how this is possible will be
focus of future work.
